---
title: "Statistical Inference Project"
author: "Guy Cole"
date: "October 2, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Simulation Exercise

## Overview

Investigate the exponential distribution in R and compare it w/the Central Limit Theorem.  

As mandated by assignment, exponential distribution will be simulated in R using rexp(n, lambda) where lambda is the rate parameter.  Distribution mean and standard deviation are 1/lambda and lambda = 0.2 for all simulations.

Assigment mandates investigation of "distribution of averages of 40 exponentials" w/1000 simulations.

## Simulations

Initialization.

```{r simulation1}
set.seed(54321)
lambda <- 0.2
n <- 40
simulationLimit <- 1000
standardDeviation <- 1/lambda
```

## Sample Mean vs Theoretical Mean

```{r sampleMeanVsTheoryMean}
simulation <- rexp(simulationLimit * n, rate = lambda)

simulatedMean <- mean(simulation)
simulatedMean

theoreticalMean <- 1/lambda
theoreticalMean
```

## Sample Variance vs Theoretical Variance

```{r sampleVarianceVsTheoryVariance1}
datum <- NULL
for (ii in 1:simulationLimit) {
  datum <- c(datum, mean(rexp(n, lambda)))
}
```

### Illustrate sample mean vs theoretical mean.  

```{r sampleVarianceVsTheoryVariance2}
library(ggplot2)

dataFrame1 <- data.frame(header = c("Sample Mean", "Theoretical Mean"), values = c(simulatedMean, theoreticalMean))

gg <- ggplot(NULL, aes(x=datum))

plot1 <- gg + geom_histogram(aes(y=..density..), color="blue", fill="red", binwidth=0.1) + labs(title = "Sample vs Theoretical Mean") + geom_density(color="black") + geom_vline(data=dataFrame1, aes(xintercept=values, linetype=header, color=header), show.legend=TRUE)

plot1
```

Standard deviation

```{r sampleVarianceVsTheoryVariance3}
sd(datum)
```

Variance

```{r sampleVarianceVsTheoryVariance4}
varianceDatum <- sd(datum)^2
varianceDatum
```

Theoretical variance

```{r sampleVarianceVsTheoryVariance5}
theoreticalDatum <- (1/lambda)^2/n
theoreticalDatum
```

## Normal Distribution

The distribution plot (above) clearly exhibits gaussian bell curve shape, indicated normal distribution.  Plot below uses qqnorm() to compare simulation to normal, indicating close approximation.

```{r normalDist1}
qqnorm(datum)
qqline(datum, col=10)
```

# Tooth Growth

## Overview

Explore the R datasets "ToothGrowth" data.

"ToothGrowth" captures the length of teeth in ten guinea pigs w/three dose levels of Vitamin C (0.5, 1 and 2 mg), delivered via orange juice or ascorbic acid.


```{r toothGrow}
library(dplyr)
library(knitr)
library(datasets)

data("ToothGrowth")
```

Data Dictionary

* [,1] len numeric Tooth length
* [,2] supp factor Supplement type (VC or OJ).
* [,3] dose numeric Dose in milligrams.

```{r toothGrow2}
str(ToothGrowth)
head(ToothGrowth)
summary(ToothGrowth)
```

## Initial Analysis

Initial analysis indicates that longer tooth length correlates to higher vitamin C dosage.

```{r toothGrow3}
summaryDataFrame <- ToothGrowth %>% group_by(supp, dose) %>% summarise(count = n(), mean=mean(len), median=median(len), "standard deviation" = sd(len))

kable (summaryDataFrame ,digits = 3, align = 'c')
```

```{r toothGrow4}
ggplot(data=ToothGrowth, aes(x=supp, y=len))+
        geom_boxplot(aes(fill=supp))+
        xlab("supplement type") +
        ylab("tooth length")+ 
        ggtitle("tooth length by supplement type ")
```

## Confidence Interval Testing

Employ t test to verify null hypotheses (both supply menthods produce same results).

```{r toothGrow5}
t.test(ToothGrowth$len[ToothGrowth$supp=="OJ"], ToothGrowth$len[ToothGrowth$supp=="VC"], paired=FALSE, var.equal=FALSE)
```

Return p-value of 0.06063 which is not large enough to reject null hypothesis.  Conclusion is that tooth length not caused by supply method.

Test increasing dose from 0.5 to 1.

```{r toothGrow6}
t.test(ToothGrowth$len[ToothGrowth$dose==0.5], ToothGrowth$len[ToothGrowth$dose==1], paired=FALSE, var.equal=TRUE)
```

Dramatically lower p value (1.266e-07) implies dosage impacts mean.

```{r toothGrow7}
t.test(ToothGrowth$len[ToothGrowth$dose==1], ToothGrowth$len[ToothGrowth$dose==2], paired=FALSE, var.equal=TRUE)
```

Which also yields a very low p value (1.811e-05) which implies that increased vitamin C dosage encourages tooth growth.

## Conclusion

Analysis using t-tests indicate that delivery methods do not impact tooth growth.  Increased levels of vitamin C dosage encourages increased tooth growth.

